---
title: "crab_data_processing_2018"
author: "Rebecca Cates"
date: "5/18/2020"
output: html_document
---
Brief description of data:
Crab data abundance data was collected along the western coastline of Prince of Wales island from July to August, 2018 as part of a project studying eelgrass ecosystems. Four strings with 2 pots each were were set at 6 different sites. Pots soaked for ~24 hours before being pulled up to identify, measure and sex the contents of the pot. This dataset is part of a time series that spans the years 2017-2020. 

Goal of this script :
1) Rename columns to fix errors and match universal naming scheme defined by APECS lab. 
2) Check maximum and minimum values of numeric values to see if they make sense within the context of      the column.
3) Trim whitespace in cells.
4) Check that absent data is referenced using the appropriate null value ("NA").
5) Check for accuracy of data type and fix if needed (e.g., factor, numeric, character)
6) Check for consistency across sites and taxonomy using APECS lab universal naming conventions
7) Create a primary key column combining species code and year so that data from 2017, 2018 and 2019 can    be joined eventually
8) Create a secondary key column combining year and site to eventually join crab data with sea otter       data based on site
9) Check for dupilcates of unique values and fix if needed
10) Reorganize columns to tidy appearance
11) Change name of Metacarcinus magister and Metacarcinus gracilis to "Metacarcinus spp." because of inconsistency in identification in the field

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load packages
```{r warning=FALSE}
library(dplyr)
library(tidyr)
```

#Read in raw raw_crab18 data and replace blanks with NA
#Read in naming conventions for sites and species from KNB
```{r}
raw_crab18<- read.csv("C:/Users/cates/OneDrive/Documents/UAF/Data 2019/Crab/raw_crab_data_2018.csv", header = TRUE, stringsAsFactors = FALSE, na.strings=c("","NA")) #eventually source from KNB

site_names <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Ac9c99ce9-fbdd-4879-a2c9-c90448cdba7b", method = "libcurl"),stringsAsFactors = FALSE)


sp_names <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3A08b8bcd1-3451-4a3a-a9ee-f3ad8b1a437e", method = "libcurl"),stringsAsFactors = FALSE)

```

#Rename columns
```{r}
#rename columns in raw_crab18 to match columns in sp_names and site_names
#necessary for left_join later
raw_crab18 <- raw_crab18 %>% 
  rename(site_2018 = ï..site) %>% 
  rename(Sp_ScientificName = species_scientific)%>% 
  rename(SpCode = sp_code)%>% 
  rename(Sp_CommonName = species_common)%>% 
  rename(Taxa = taxon)

```

#Add APECS lab universal site naming convention column $bay_code and $bay_sample
```{r}

#create temporary site_names data frame with columns of interest to be joined with raw_crab2017 dataframe
site_names.tmp = site_names[,c("site_2018", "bay_code", "bay_sample")]
colnames(site_names.tmp)

#join site_names.tmp and raw_crab18
raw_crab18 = left_join(raw_crab18, site_names.tmp, by = "site_2018")
head(raw_crab18)

```

#Check spelling in raw_crab18$Sp_ScientificName, $SpCode, Sp_CommonName
```{r}
#Check for spelling errors in raw_crab18$Sp_ScientificName
table(raw_crab18$Sp_ScientificName)
#2 mistakes
#Let's fix em
raw_crab18$Sp_ScientificName[raw_crab18$Sp_ScientificName == "Metacarcinus magister \n"]  = "Metacarcinus magister"
raw_crab18$Sp_ScientificName[raw_crab18$Sp_ScientificName == "Metacarcinus gracillis"]  = "Metacarcinus gracilis"
raw_crab18$Sp_ScientificName[raw_crab18$Sp_ScientificName == "Lepidopsetta bilineata"]  = "Lepidopsetta spp."
raw_crab18$Sp_ScientificName[raw_crab18$Sp_ScientificName == "Pycnopodia helianthodes"]  = "Pycnopodia helianthoides"


#Check for spelling errors in raw_crab18$SpCode
table(raw_crab18$SpCode)#5 mistakes
#Let's fix em
raw_crab18$SpCode[raw_crab18$SpCode == "SCULSTG"]  = "SCULPSTG"
raw_crab18$SpCode[raw_crab18$SpCode == "CRABGRC"]  = "CRABGRA"
raw_crab18$SpCode[raw_crab18$SpCode == "SOLERCK"]  = "SOLEROC"
raw_crab18$SpCode[raw_crab18$SpCode == "STARSF"]  = "STARPYCN"
raw_crab18$SpCode[raw_crab18$SpCode == "UNIDFLT"]  = "UNFLAT"


#Check for spelling errors in raw_crab18$Sp_CommonName
table(raw_crab18$Sp_CommonName) # 4 mistakes
#Let's fix em
raw_crab18$Sp_CommonName[raw_crab18$Sp_CommonName == "Sunflower star"]  = "Sunflower sea star"
raw_crab18$Sp_CommonName[raw_crab18$Sp_CommonName == "Unknown flatfish"]  = "Juvenile flatfish"
raw_crab18$Sp_CommonName[raw_crab18$Sp_CommonName == "Rock Sole"]  = "Rock sole"
raw_crab18$Sp_CommonName[raw_crab18$Sp_CommonName == "Graceful crab"]  = "Graceful rock crab"

```

#add missing species in sp_names
#These errors should be fixed in the KNB file itself when we have permission to edit and this code section can be deleted
```{r}
sp_names <- add_row(sp_names, SpCode = "STARGP", Sp_CommonName = "Giant pink sea star", Sp_ScientificName = "Pisaster brevispinus", Taxa = "Echinodermata") 


```

#Join sp_names and raw_crab18 and check for mismatches
```{r}
#join sp_names and raw_crab18
raw_crab18 = left_join(raw_crab18, sp_names, by = "SpCode")
head(raw_crab18)

#Check that duplicate columns are matching
#Sp_CommonName
table(raw_crab18$Sp_CommonName.x != raw_crab18$Sp_CommonName.y) #1 mismatched
which(raw_crab18$Sp_CommonName.x != raw_crab18$Sp_CommonName.y) #row 108
raw_crab18$Sp_CommonName.y = trimws(raw_crab18$Sp_CommonName.y, which = c("both", "left", "right"))
table(raw_crab18$Sp_CommonName.x != raw_crab18$Sp_CommonName.y) #Looks good

#Sp_ScientificName
table(raw_crab18$Sp_ScientificName.x != raw_crab18$Sp_ScientificName.y) #Looks good

#Taxa
table(raw_crab18$Taxa.x != raw_crab18$Taxa.y) 
#Different systems were used for $Taxa.x and $Taxa.y so none match. I am keeping $Taxa.y in the cleaned dataset
    
```

#Rename columns, get rid of duplicate columns and reorder
```{r}
#remove duplicate columns
raw_crab18 <- raw_crab18 %>% 
  select(-site_2018, -Taxa.x, -Sp_CommonName.y,-Sp_ScientificName.y)

#check
colnames(raw_crab18)

#rename the remaining duplicated columns
raw_crab18 <- raw_crab18 %>% 
  rename(Sp_ScientificName = Sp_ScientificName.x) %>% 
  rename(Sp_CommonName = Sp_CommonName.x) %>% 
  rename(Taxa = Taxa.y)

#reorder to make tidy
raw_crab18 <- raw_crab18 %>% 
  select(bay_code, bay_sample, date_set, YYYYMMDD_set, time_set, date_pulled, YYYYMMDD_pulled, time_pulled, time_soak_hr, trap, string, Sp_CommonName, Sp_ScientificName, SpCode, Taxa, carapace_width_mm, sex, fork_length_mm, unmeasured, notes)

#check
colnames(raw_crab18)

```

#Rename Metacarcinus gracilis and magister to Metacarcinus spp. to account for identification inconsistency in field
```{r}
raw_crab18$Sp_ScientificName[raw_crab18$Sp_ScientificName == "Metacarcinus magister"] = "Metacarcinus spp."
raw_crab18$Sp_ScientificName[raw_crab18$Sp_ScientificName == "Metacarcinus gracilis"] = "Metacarcinus spp."

raw_crab18$SpCode[raw_crab18$SpCode == "CRABDUN"] = "CRABMET"
raw_crab18$SpCode[raw_crab18$SpCode == "CRABGRA"] = "CRABMET"

raw_crab18$Sp_CommonName[raw_crab18$Sp_CommonName == "Dungeness crab"] = "Metacarcinus spp."
raw_crab18$Sp_CommonName[raw_crab18$Sp_CommonName == "Graceful rock crab"] = "Metacarcinus spp."

```

#Create a primary key column combining species code and year to join all years together at later point
```{r}

raw_crab18$pri.key = NA

raw_crab18$pri.key <- paste(raw_crab18$SpCode, "18", sep = ".")

```

#Create a secondary key column combining year and site to join crab data with sea otter data based on site
```{r}
raw_crab18$sec_key = NA

raw_crab18$sec_key <- paste(raw_crab18$bay_code, "18", sep = ".")
```

#Trim whitespace

```{r}
#remove leading and trailing whitespace
raw_crab18 <- raw_crab18 %>%
  mutate_if(is.character, trimws)

```

#Convert columns to factors that are characters
```{r}
raw_crab18 <- raw_crab18 %>%
  mutate_if(is.character,as.factor)

str(raw_crab18)

```


#Check number of unique levels within columns that are factors and fix if necessary

```{r}
str(raw_crab18)

#Check for duplicates
length(table(raw_crab18$bay_code)) #Should be 6 levels and there are
length(unique(raw_crab18$bay_code)) #6 unique levels, no duplicates

length(table(raw_crab18$trap)) #Should be 2 levels and there are
length(unique(raw_crab18$trap))#2 unique levels, no duplicates

length(table(raw_crab18$sex)) #Should be 2 levels and there are
length(unique(raw_crab18$sex)) #2 unique levels, no duplicates

length(table(raw_crab18$Taxa)) #Should be 3 levels and there are
length(unique(raw_crab18$Taxa)) #3 unique levels, no duplicates

#Check to see if unique lengths of $Sp_ScientificName, $SpCode, $SpCommonName match
length(unique(raw_crab18$Sp_ScientificName)) == length(unique(raw_crab18$Sp_CommonName)) 
length(unique(raw_crab18$Sp_ScientificName))== length(unique(raw_crab18$SpCode))
#Looks good

```

#Check for NA values
```{r}
#Summarize NA values over all columns
raw_crab18 %>%
  select(everything()) %>%
  summarise_all(funs(sum(is.na(.))))
#NAs only in columns where they should be

```

#Check maximum and minimum values of numeric values to see if they make sense within context
```{r}
summary(raw_crab18)
```

#write clean dataframe to csv
```{r}
write.csv(raw_crab18, "clean_crab_data_2018.csv", row.names = FALSE)
```

